---
title: "Tree Equity Student Presentation"
author: "Your Name Here"
format: 
  revealjs:
    slide-number: true
    theme: simple
    transition: slide
    scrollable: true
    center: true
    toc: false
    embed-resources: true
    echo: true
---

## Instructions for Students (Delete This Slide Before Submitting)

- This is your **template** for presenting your project.

- Your final presentation should have **4 slides**:
  1. Research Question + Dataset
  2. Data Wrangling Code
  3. Plot + ggplot Code
  4. Key Insights

Once you've completed the slides, **delete this instruction slide** before rendering.

Note that the datasets are here: https://www.treeequityscore.org/methodology?tab=data-download

The data glossary is here: https://www.treeequityscore.org/methodology?tab=data-glossary

## Slide 1: Dataset & Research Question

**Dataset Chosen:**  

> e.g., Tree Equity data from Alabama, Arizona, etc.

> Include the source of the data (link + resource name)

```{r}
#| echo: false

# read in your dataset here
# see how the data were loaded in the wrangling and ggplot examples
```

**Research Question:**  

> The question you are investigating using the TES dataset.



## Slide 2: Data Wrangling

Use tidyverse functions like `filter()`, `mutate()`, `select()`, `group_by()`, `summarise()`, etc. Show all the code that you used to clean or transform the data for your plot.

```{r}


```

## Slide 3: Plot

Use **ggplot2** to visualize your findings. Include code and plot.

```{r}


```


## Slide 4: Key Takeaways

What did you learn from your analysis?
Your answer should make statements about what you learned, so you should delete all of this text and write down your takeaways.
For example, you might respond to some / all of these questions:

- What patterns or disparities did you observe?
- What conclusions can you draw from your plot?
- Why do these findings matter?

