---
title: "CA Public Transportation"
author: "Paityn Richardson"
format:
  revealjs:
    theme: simple
    transition: fade
    incremental: true
    scrollable: true
execute:
  warning: false
  message: false
---

```{r}
#: include: false
options(pillar.width = 70)
```

## Public Transportation in Urbn Planning

**Public Transit is a Key factor in shaping how a city/society functions **

  - Reduces Traffic Congestion and Pollution (sustainability)
  - Increases access to Jobs, Education, Healthcare, etc.
  - Impacts Housing affordability and community equity


## Public Transportation in Urbn Planning

### Useful Analyses include:

- Trends in ridership over time
- Efficiency: Cost per rider or mile
- Equity: *Who uses transit?* *Where is/isn't transit accessible?*


- *What types of variables should we look for in a dataset/sets to conduct these useful analyses?*


## Dataset: National Transit Database (NTD)

- Data is filtered from US data -> CA specific Data

. . .


```{r}
#| echo: false
#| eval: true
library(readr)
library(tidyverse)

ca_transit <- read_csv("ca_transit.csv")

head(ca_transit, 1)
```

. . . 

**Discuss:**

- *What are the variables? *
- *What kinds of questions can we ask?*

## Wrangling Example

> Goal: Compare Agencies Ridership over time

. . .

```{r}
#| echo: true
#| eval: true 
#| code-line-numbers: 1|2|3|4

summary <- ca_transit|>
  group_by(Agency, Year)|>
  summarise(tot_pass = sum(Unlinked.Passenger.Trips), .groups = "drop")|>
  arrange(desc(tot_pass))

```

. . . 

*What does `summary` look like?*


## Visualizing with ggplot2

> Let's look at the agencies with the top 5 `tot_pass`

. . . 

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: 1|2|3|4|6|7|8|9|10|11|12

top5 <- summary|>
  filter(Year == 2023)|>
  slice_max(tot_pass, n = 10)|>
  pull(Agency)

summary |>
  filter(Agency %in% top5) |>
  ggplot(aes(x = Year, y = tot_pass, color = Agency)) +
  geom_line(linewidth = .6) +
  labs(title = "Transit Ridership Over Time (Top 5 Agencies)",
       y = "Persons",
       x = "Year")
```

## 

```{r}
#| echo: false
#| eval: true
#| code-line-numbers: 1|2|3|4|6|7|8|9|10|11|12

top5 <- summary|>
  filter(Year == 2023)|>
  slice_max(tot_pass, n = 10)|>
  pull(Agency)

summary |>
  filter(Agency %in% top5) |>
  ggplot(aes(x = Year, y = tot_pass, color = Agency)) +
  geom_line(linewidth = .6) +
  labs(title = "Transit Ridership Over Time (Top 5 Agencies)",
       y = "Persons",
       x = "Year")
```

## Your Turn

**Work With a partner/group to:**

  - Filter data for specific Agency -> "Pomona Valley Transit Authority"
  - Create line plot of `tot_pass` over time

  
##


```{r}
#| eval: true
#| echo: true

summary |> 
  filter(Agency == "Pomona Valley Transportation Authority") |>
  ggplot(aes(x = Year, y = tot_pass)) +
  geom_line(linewidth = 0.6) +
  labs(
    title = "Pomona Valley Transportation Authority Ridership Over Time",
    y = "Riders/(pop.)",
    x = "Year"
  )
```


## New Dataset: California Transit Stops

- Each **row** is a public transit stop in California.

. . . 

```{r}
#| echo: false
#| eval: true
#| include: false
library(sf)
stop_data <- st_read("output_shapefile/CA_Transit_Stops.shp")

```

```{r}
#| echo: false
head(stop_data, 5) |> 
  select(-district_n) |> 
  as_tibble() |> 
  print(width = Inf)
```

## Variables

**Similar to previous dataset:**
 
  - Agency

. . . 

**New:**

  - stop_name/id
  - routetype
  - **geomotry?**
  
. . .

This dataset is stored as a **shapefile**.

## What Is a Shapefile?

A **shapefile** is a special type of dataset used to:

- Store geographic information
- Plot points, lines, or areas on a map
- Perform spatial joins and analysis

```{r}
#| eval: true
#| echo: true
#| include: false

geomotry <- stop_data|>
  select(GEOID, geometry)

head(geomotry, 5)
```

## What Can this Data Tell Us?

**On its own, this data shows:** 

  - Where stops are located
  - How many stops an agency operates
  - How often stop is used/How many routes serve a stop

. . . 


**What the data doesn't show is:**

  - Who lives near a certain stop
  - What the area is like demographically
  - Whether transit access is equitable
  - What causes someone to take transit
  
## Adding Context

. . . 

> A common data wrangling practice used to connect datasets that share a common variables is called **joining**

- We can use functions like `right_join()` or `left_join()` to connect our `stop_data` to a dataset with **socioeconomic** identifiers. 


## Adding Context

```{r}
#| echo: true
#| eval: true
#| include: false

census_data <- st_read("output_shapefile2/CA_Census_Data.shp")
```

- Each stop in our `stop_data` is connected to a `GEOID`
- In our census data, each `GEOID` has information on the areas demographics and socioeconomic indicators

. . . 

> This means we can join these 2 datasets on the variable `GEOID` !

## Right Join vs. Left Join

> Let's Compare our resulting datasets after using the functions `left_join()` and `right_join()`

##

```{r}
#| echo: true

right_joined <- census_data |> 
  right_join(stop_data |> st_drop_geometry(), by = "GEOID")

right_joined

```

##

```{r}
#| echo: true

left_joined <- census_data |> 
  left_join(stop_data|> st_drop_geometry(), by = "GEOID")

left_joined
```

## Right Join vs. Left Join

- `left_join()`: Keep **all rows from the left** (`census_data`), attach matching rows from right
- `right_join()`: Keep **all rows from the right** (`stop_data`), attach matching left values

. . .

➡️ Use `left_join()` if you want **every census area**, even if it has no stop 

➡️ Use `right_join()` if you want **every transit stop** 

##

> We want to keep every transit stop,even in GEOIDS with multiple stops, meaning we must use right_join()

. . .


```{r}
#| echo: true
#| code-line-numbers: 1|2

full_data <- census_data |> 
  right_join(stop_data |> st_drop_geometry(), by = "GEOID") 

full_data

```

## Exploring: Poverty & Transit Access 

Now that we have out datasets combined, we can start exploring the relationships between our stop data and our census data. 

. . .


> **Q: How does `pctpov` (percent in poverty) relate to the number of transit stops in a neighborhood?**

## Exploring: Poverty & Transit Access 

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: 1|2|3|5|6|7|8|9|10|11|12|13

full_data <- full_data |>
  group_by(GEOID)|>
  summarize(stop_count = n(),
            pctpov = first(pctpov),
            pctpoc = first(pctpoc),
            mdn_ncm = first(mdn_ncm),
            pct_w_c = first(pct_w_c))


full_data|>
  ggplot(aes(x = pctpov, y = stop_count)) +
  geom_point(color = "lightgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Transit Stops vs Poverty Rate",
       x = "Percent in Poverty",
       y = "Number of Transit Stops") + 
  theme_grey()

```

##

```{r}
#| echo: false
#| eval: true
#| code-line-numbers: 1|2|3|5|6|7|8|9|10|11|12|13

full_data <- full_data |>
  group_by(GEOID)|>
  mutate(stop_count = n())


full_data|>
  ggplot(aes(x = pctpov, y = stop_count)) +
  geom_point(color = "lightgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Transit Stops vs Poverty Rate",
       x = "Percent in Poverty",
       y = "Number of Transit Stops") + 
  theme_grey()

```

## Exploring: Median Income & Transit Access

> **Q: Do higher-income areas have more or fewer transit stops?**

. . .


```{r}
#| echo: true
#| eval: false
#| code-line-numbers: 1|2|3|4|5|6|7|8

full_data|>
  ggplot(aes(x = mdn_ncm, y = stop_count)) + 
  geom_point(color = "lightblue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Transit Stops vs Median Income",
       x = "Median Income",
       y = "Number of Transit Stops") + 
  theme_grey()
```

##

```{r}
#| echo: false
#| eval: true
#| code-line-numbers: 1|2|3|4|5|6|7|8

full_data|>
  ggplot(aes(x = mdn_ncm, y = stop_count)) + 
  geom_point(color = "lightblue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Transit Stops vs Median Income",
       x = "Median Income",
       y = "Number of Transit Stops") + 
  theme_grey()
```

## One more!

> **Q: Do areas where more people don’t own cars have better transit coverage?**

. . .


```{r}
#| echo: true
#| eval: false
#| code-line-numbers: 1|2|3|4|5|6|7|8

full_data|>
  ggplot(aes(x = pct_w_c, y = stop_count)) + 
  geom_point(color = "yellow") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Transit Stops vs Car Ownership",
       x = "Percent Without",
       y = "Number of Transit Stops") + 
  theme_grey()
```

##

```{r}
#| echo: false
#| eval: true
#| code-line-numbers: 1|2|3|4|5|6|7|8

full_data|>
  ggplot(aes(x = pct_w_c, y = stop_count)) + 
  geom_point(color = "yellow") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Transit Stops vs Car Ownership",
       x = "Percent Without",
       y = "Number of Transit Stops") + 
  theme_grey()
```

## Exploration Questions

> These plots highlight the **equity** of the CA Public Transit System. 

Why might these practices be important in areas **Outside** of CA? 

## Visualization 2.0 

With our Data we want to Show...

  - **Where** Transit Stops Are Located
  - **Which** areas are affected by socioeconomic Factors
  - **Correlation** between transit stop access and socioeconomic status
  
. . . 

> For this task, We can use a Spatial Heat Map to visualize the overlap between the two areas

## Spatial Heat Maps

- Census Data Automatically Contains the `geometry`, or placement on a map, for each GEOID
- Stop Data places the stop inside each GEOID at its (lat,long) cord.
- We can Visualize the socioeconomic factors through colors on the map 

. . .

> Our Goal: Create a Map of CA, divided into GEOID's, that compares transit stop frequency to socioeconomic factors

## 

```{r}
#| echo: false


facet_data <- full_data |>
  pivot_longer(cols = c(pctpoc, pctpov, mdn_ncm, pct_w_c),
    names_to = "census_var",
    values_to = "value") |>
  filter(!is.na(value))|>
  select(geometry, stop_count, census_var, value)

library(patchwork)

plots <- split(facet_data, facet_data$census_var) |> 
  purrr::map(~ {
    ggplot(.x) +
      geom_sf(aes(fill = value)) +
      #geom_sf_text(aes(label = stop_count), size = 0.5, color = "white") +
      scale_fill_viridis_c() +
      theme_minimal() +
      labs(title = unique(.x$census_var))
  })

wrap_plots(plots)


facet_data |> 
  filter(census_var == "pctpov") |> 
ggplot() +
      geom_sf(aes(fill = value)) +
      geom_sf_text(aes(label = stop_count), size = 0.5, color = "white") +
      scale_fill_viridis_c() +
      xlim(c(-119, -116)) + 
      ylim(c(33.5,35)) + 
      theme_minimal() 

 

```


