---
title: "Tree Equity Student Presentation"
author: "Annika Heitkemper"
format: 
  revealjs:
    slide-number: true
    theme: simple
    transition: slide
    scrollable: true
    center: true
    toc: false
    embed-resources: true
    echo: true
---


```{r}
#| echo: false
#| include: false

library(tidyverse)
tmp <- tempfile(fileext = ".zip")
or_data_zip <- download.file("https://tes-app-data-share.s3.amazonaws.com/or/or_csv.zip", 
                             tmp, mode = "wb")

unzip(tmp, list = TRUE)

or_tes <- readr::read_csv(unz(tmp, "or_tes.csv"))
```




## Slide 1: Dataset & Research Question

**Dataset Chosen:**  

> Tree Equity data from Oregon

> https://www.treeequityscore.org Tree Equity Score

```{r}
#| echo: false

# read in your dataset here
# see how the data were loaded in the wrangling and ggplot examples
```

**Research Question:**  

> What is the relationship between poverty and tree equity scores in different counties in Oregon?





## Slide 3: Plot

Use **ggplot2** to visualize your findings. Include code and plot.

```{r}
or_tes |>
  ggplot(aes(x = pctpov, y = tes, color=county)) +
  geom_point(size = 0.75) +
  facet_wrap(~county) +
  theme(legend.position="none")

```


## Slide 4: Key Takeaways


- The patterns I saw was that the bigger the population the more trees.
- The concusions I can draw is that the less poverty the more trees.
- These findings matter because we want more trees so there is less poverty.

