---
title: "Tree Equity Student Presentation"
author: "Billie Sutherland"
format: 
  revealjs:
    slide-number: true
    theme: simple
    transition: slide
    scrollable: true
    center: true
    toc: false
    embed-resources: true
    echo: true
---



```{r}
#| echo: false
#| include: false
library(tidyverse)
tmp <- tempfile(fileext = ".zip")
ca_data_zip <- download.file("https://tes-app-data-share.s3.amazonaws.com/ny/ny_csv.zip", 
                             tmp, mode = "wb")

unzip(tmp, list = TRUE)

ny_tes <- readr::read_csv(unz(tmp, "ny_tes.csv"))

```


## Slide 1: Dataset & Research Question

**Dataset Chosen:**  

> Tree Equity data from New York

> treeequityscore.org

```{r}
#| echo: false

# read in your dataset here
# see how the data were loaded in the wrangling and ggplot examples
```

**Research Question:**  

>What is the relationship between HOLC Grade and Tree Equity Score in New York?



## Slide 2: Data Wrangling

Use tidyverse functions like `filter()`, `mutate()`, `select()`, `group_by()`, `summarise()`, etc. Show all the code that you used to clean or transform the data for your plot.

```{r}
ny_tes|>
  filter(!is.na(holc_grade))

```

## Slide 3: Plot

Use **ggplot2** to visualize your findings. Include code and plot.

```{r}
ny_tes|>
  filter(!is.na(holc_grade))|>
  ggplot(aes(y=tes, x=holc_grade))+
  geom_boxplot(fill="pink")+
  geom_jitter(color="navy", alpha=0.05)

```


## Slide 4: Key Takeaways


- What patterns or disparities did you observe?

The more trees, the higher the HOLC score, but this not causal.

- What conclusions can you draw from your plot?

I can tell that places with worse HOLC scores, there are less trees.

- Why do these findings matter?

Because it is important to know that even if data sets have negative relationships, they can still be causal or not.
