---
title: "Practicing Data Wrangling with Tree Equity Data"
subtitle: "in-class worksheet viz"
author: "Paityn Richardson & Jo Hardin"
format:
  pdf: 
    fontsize: 10.5pt
    geometry: margin=0.75in
    toc: false
execute: 
  warning: false
  message: false
---


## Instructions

In this worksheet, you'll practice writing **ggplot R code** for creating data visualizations. 
Each question guides you through building layers of plots for the `ca_tes` dataset. 

**Don't stress yourself out over perfect syntax** -- the goal is to practice what you've learned about data wrangling so far.


## Practice Questions


## 1. Tree Equity vs. Poverty

> Is there a relationship between poverty (`pctpov`) and tree equity score (`tes`)?


```{r}
#| echo: true
#| eval: false

ca_tes |>
  ___(___(x = ___, y = ___)) +
  ___(color = "green") 
```


```


```

## 2. Distribution of tree equity scores

> What is the distribution of tree equity scores (`tes`)?


```{r}
#| echo: true
#| eval: false

___ |>
  ggplot(aes(___)) +
  ___(fill = "red")
```



```



```

## 3. Distribution of tree equity scores

> What is the distribution of tree equity scores (`tes`) created by a boxplot?


```{r}
#| echo: true
#| eval: false

___ |>
  ggplot(aes(___)) +
  ___(fill = "violet")
```


```



```


## 4. Tree equity vs. linguistic isolation

> Is there a relationship between linguistic isolation(`lingnorm`) and `tes` in an area. 

  - Use ggplot to plot the relationship between `tes` and `lingnorm`
  - Choose the best plot for the variable types
  - Create fitting labels for your axes and plot title



```


```



## 5. Wrangle + plot

Recreate the plot above, but summarize the information **per `county`** before plotting. That is, `Los Angeles County` should now only be represented by one dot. Summarize both `lingnorm` and `tes` using the `mean()` function.



```



```

## 6. Wrangle + plot + label

Recreate the plot above, but now add labels to the plot to indicate the county of each dot.  Use `geom_label()` or `geom_text()`. Try both! You'll need to use `label = county`...  do you need `aes()`? 



```


```


## 7. Wrangle + plot + label

There is an R package called **ggrepel**.  It includes a geom called `geom_label_repel`. Try it!  (To use a function from a package, you can write: `ggrepel::geom_label_repel()`.)

**Hint:** If R can't find the **ggrepel** package, you may need to type `install.packages("ggrepel")` in the Console down below.


```


```



